---
title: "Exploring Stub Model Object Structures"
subtitle: "rpart, gbm, xgboost — edges, nodes, and implementation notes"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Purpose

Before implementing `edgelist()` / `nodelist()` for rpart, gbm, and xgboost, we need
to understand exactly how each object stores its tree structure. Getting this wrong is
easy — Claude initially misunderstood the `tree` package format, so we're being
careful here.

Each section:

1. Fits a model with enough depth to exercise the structure meaningfully
2. Inspects the raw internal structure
3. Interprets what constitutes "edges" and "nodes"
4. Proposes an implementation approach (no code)
5. Describes a validation strategy
6. Flags complexity / gotchas


---

# rpart

## Fit a model and inspect

We use `kyphosis` (rpart's built-in dataset) — it has a factor response, a mix of
predictors, and grows a reasonably deep tree. We also lower `cp` to let the tree branch
out further so we can see non-consecutive node IDs (gaps from pruning/heap numbering).

```{r rpart-fit}
library(rpart)

# kyphosis: 81 obs, 4 vars (Age, Number, Start — numeric; Kyphosis — factor response)
# Low cp + modest minsplit to get a deeper tree
fit_rpart <- rpart(Kyphosis ~ ., data = kyphosis,
                   control = rpart.control(cp = 0.005, minsplit = 5, maxdepth = 5))
cat("Number of nodes:", nrow(fit_rpart$frame), "\n")
cat("Number of leaves:", sum(fit_rpart$frame$var == "<leaf>"), "\n")
cat("Number of internal:", sum(fit_rpart$frame$var != "<leaf>"), "\n")
```

### The `frame` component — one row per node

```{r rpart-frame}
fit_rpart$frame
```

Row names are node IDs. `var == "<leaf>"` marks terminal nodes.

### Node ID numbering scheme

rpart uses **binary heap numbering** (1-indexed):

- Root = 1
- Left child of node k = 2k
- Right child of node k = 2k + 1
- Parent of node k = k %/% 2

```{r rpart-node-ids}
node_ids <- as.integer(rownames(fit_rpart$frame))
cat("Node IDs:", node_ids, "\n")
cat("Min:", min(node_ids), " Max:", max(node_ids), "\n")
cat("Count:", length(node_ids), "\n")
cat("Are they consecutive?", identical(node_ids, seq_along(node_ids)), "\n")
cat("Gaps show where pruned branches would have been\n")

# Verify parent-child relationships
child_ids <- node_ids[node_ids != 1L]
parent_ids <- child_ids %/% 2L
cat("\nEdges (parent -> child):\n")
edges_rpart <- data.frame(
  from = parent_ids, to = child_ids,
  direction = ifelse(child_ids %% 2 == 0, "left", "right"))
print(edges_rpart)

# Verify every parent actually exists in the node set
cat("\nAll parents in node set?", all(parent_ids %in% node_ids), "\n")
```

### The `splits` matrix

```{r rpart-splits}
cat("Dimensions:", dim(fit_rpart$splits), "\n\n")
fit_rpart$splits
```

Rows map to frame rows **sequentially** (consuming 1 + ncompete + nsurrogate rows per
internal node). The first consumed row is the primary split.

- `ncat < 0` = continuous split; `index` is the cutpoint
- `ncat > 1` = categorical split; `index` is a row into `csplit`

```{r rpart-splits-mapping}
# Show how splits rows map to frame rows
mapping <- data.frame(
  node = as.integer(rownames(fit_rpart$frame)),
  var = fit_rpart$frame$var,
  ncompete = fit_rpart$frame$ncompete,
  nsurrogate = fit_rpart$frame$nsurrogate,
  rows_consumed = ifelse(
    fit_rpart$frame$var == "<leaf>", 0,
    1 + fit_rpart$frame$ncompete + fit_rpart$frame$nsurrogate
  )
)
print(mapping)
cat("\nTotal splits rows consumed:", sum(mapping$rows_consumed), "\n")
cat("Actual splits rows:", nrow(fit_rpart$splits), "\n")
```

### Walk the splits matrix to extract primary split per node

This is the algorithm we'll need in the implementation — mapping each internal node
to its primary split threshold.

```{r rpart-splits-walk}
node_ids <- as.integer(rownames(fit_rpart$frame))
is_leaf <- fit_rpart$frame$var == "<leaf>"
primary_split <- data.frame(
  node = integer(0), var = character(0),
  ncat = numeric(0), index = numeric(0))

split_row <- 1L
for (i in seq_along(node_ids)) {
  if (is_leaf[i]) next
  primary_split <- rbind(primary_split, data.frame(
    node = node_ids[i],
    var = rownames(fit_rpart$splits)[split_row],
    ncat = fit_rpart$splits[split_row, "ncat"],
    index = fit_rpart$splits[split_row, "index"]
  ))
  n_consume <- 1L + fit_rpart$frame$ncompete[i] + fit_rpart$frame$nsurrogate[i]
  split_row <- split_row + n_consume
}
cat("Primary split per internal node:\n")
print(primary_split)
cat("\n(ncat < 0 = continuous, ncat > 1 = categorical)\n")
```

### Edge labels via `labels()`

```{r rpart-labels}
edge_labels <- labels(fit_rpart, collapse = TRUE)
data.frame(node = node_ids, label = edge_labels)
```

Element i is the condition on the edge *leading into* node i. Root gets `"root"`.

### GOTCHA: `ncat` sign controls split direction

For continuous splits, `ncat` can be -1 or +1. The sign determines whether `<` or `>=`
goes to the left child. **Do not assume left always means `<`.**

```{r rpart-ncat-direction}
# Show the ncat sign and resulting label direction for each internal node
cat("ncat sign -> label direction per internal node:\n\n")
child_ids <- node_ids[node_ids != 1L]
parent_ids <- child_ids %/% 2L

lbl_collapse <- labels(fit_rpart, collapse = TRUE)
names(lbl_collapse) <- as.character(node_ids)

lbl_matrix <- labels(fit_rpart, collapse = FALSE)
rownames(lbl_matrix) <- as.character(node_ids)

for (r in seq_len(nrow(primary_split))) {
  nid <- primary_split$node[r]
  ncat_val <- primary_split$ncat[r]
  idx_val <- primary_split$index[r]
  left_child <- 2 * nid
  right_child <- 2 * nid + 1

  cat(sprintf("Node %d: %s, ncat=%+d, cutpoint=%.1f\n",
              nid, primary_split$var[r], ncat_val, idx_val))
  if (left_child %in% node_ids) {
    cat(sprintf("  LEFT  child %d: \"%s\"\n", left_child,
                lbl_collapse[as.character(left_child)]))
  }
  if (right_child %in% node_ids) {
    cat(sprintf("  RIGHT child %d: \"%s\"\n", right_child,
                lbl_collapse[as.character(right_child)]))
  }
  expected_left_op <- if (ncat_val < 0) "<" else ">="
  cat(sprintf("  -> ncat=%+d means left gets \"%s\", right gets \"%s\"\n\n",
              ncat_val, expected_left_op,
              if (expected_left_op == "<") ">=" else "<"))
}
```

### Visualize for validation

```{r rpart-plot, fig.width=10, fig.height=7}
plot(fit_rpart, uniform = TRUE, margin = 0.1)
text(fit_rpart, use.n = TRUE, cex = 0.7)
title(paste("rpart tree —", nrow(fit_rpart$frame), "nodes,",
            nrow(edges_rpart), "edges"))
```

Count the nodes and edges in the plot and verify they match our extraction above.

### `path.rpart()` — alternative validation

```{r rpart-path}
# path.rpart returns the root-to-node path for each node
paths <- path.rpart(fit_rpart, node = node_ids, print.it = FALSE)
cat("Path to deepest node (", names(paths)[length(paths)], "):\n")
print(paths[[length(paths)]])
cat("\nPath length should equal depth + 1\n")
```

### `csplit` — categorical split directions (if present)

```{r rpart-csplit}
if (!is.null(fit_rpart$csplit)) {
  cat("csplit matrix present (", nrow(fit_rpart$csplit), "rows):\n")
  print(fit_rpart$csplit)
  cat("\nValues: 1=LEFT, 2=missing, 3=RIGHT\n")
} else {
  cat("No csplit matrix (kyphosis uses only numeric predictors)\n")
}
```

### Test with categorical predictors — use a mixed dataset

```{r rpart-categorical}
# Build a dataset with both numeric and categorical predictors
set.seed(123)
n <- 200
df_mixed <- data.frame(
  y = factor(sample(c("A","B","C"), n, replace = TRUE,
                    prob = c(0.5, 0.3, 0.2))),
  color = factor(sample(c("red","green","blue","yellow"), n, replace = TRUE)),
  size = factor(sample(c("S","M","L","XL"), n, replace = TRUE)),
  score = rnorm(n),
  count = rpois(n, 5)
)
# Make y partially predictable from predictors
df_mixed$y[df_mixed$color == "red" & df_mixed$score > 0] <- "A"
df_mixed$y[df_mixed$size == "XL"] <- "C"
df_mixed$y <- factor(df_mixed$y)

fit_cat <- rpart(y ~ ., data = df_mixed,
                 control = rpart.control(cp = 0.005, minsplit = 5, maxdepth = 5))
cat("Nodes:", nrow(fit_cat$frame), "\n")
cat("Leaves:", sum(fit_cat$frame$var == "<leaf>"), "\n\n")

if (!is.null(fit_cat$csplit)) {
  cat("csplit matrix (", nrow(fit_cat$csplit), "rows x",
      ncol(fit_cat$csplit), "cols):\n")
  print(fit_cat$csplit)
  cat("\nValues: 1=LEFT, 2=missing/unused, 3=RIGHT\n")
  cat("\nxlevels:\n")
  str(attr(fit_cat, "xlevels"))
  cat("\nEdge labels for categorical tree:\n")
  print(labels(fit_cat, collapse = TRUE))
} else {
  cat("(no categorical splits found — try a different seed)\n")
}
```

```{r rpart-cat-plot, fig.width=10, fig.height=7}
if (nrow(fit_cat$frame) > 1) {
  plot(fit_cat, uniform = TRUE, margin = 0.1)
  text(fit_cat, use.n = TRUE, cex = 0.7)
  title("rpart with categorical + numeric predictors")
}
```


## My interpretation of edges and nodes

**Nodes** = rows of `frame`. Each has an ID (from rownames), a split variable (or
`"<leaf>"`), observation count, deviance, and predicted value.

**Edges** = parent-child relationships derived purely from node IDs via the binary heap
rule. Every non-root node k has parent k %/% 2. Edge labels come from `labels(fit)`.

**This is structurally simple** — the parent-child relationship is implicit in the
numbering, not stored in explicit "left_daughter"/"right_daughter" columns like
randomForest. No traversal algorithm needed.

**IMPORTANT: the `ncat` sign controls split direction.** For continuous variables,
`ncat` can be -1 or +1:

- `ncat < 0`, left (even) child: `x < cutpoint`
- `ncat < 0`, right (odd) child: `x >= cutpoint`
- `ncat > 0`, left (even) child: `x >= cutpoint`
- `ncat > 0`, right (odd) child: `x < cutpoint`

This means left children do NOT always get `<`. The direction depends on the specific
split. `labels(fit, collapse = TRUE)` handles this correctly, so using it for edge
labels is the safest approach. If we want to extract `split_op` ourselves, we must check
the `ncat` sign from the splits matrix.

## Approximate implementation approach

**`edgelist.rpart`:**

1. Extract `node_ids` from `rownames(frame)`
2. For non-root nodes: `parent = id %/% 2`, `direction = "left"/"right"` from even/odd
3. Get edge labels from `labels(fit, collapse = TRUE)` — this already accounts for
   `ncat` sign, so it's the safest source for split conditions
4. Walk the `splits` matrix sequentially to extract primary split variable/threshold per
   internal node (for the `split_var`, `split_point` columns)
5. For `split_op`: derive from `ncat` sign and child direction (see truth table above),
   OR parse from the label string
6. Handle categorical splits via `csplit` matrix when `ncat > 1`
7. Edge case: stump (single node, no splits) -> empty edgelist

**`nodelist.rpart`:**

1. One row per node from `frame`
2. Include: name (node ID), var, n, dev, yval, is_leaf, label
3. For classification: decode `yval` to class name via `attr(fit, "ylevels")`
4. Could optionally include depth via `floor(log2(node_id))`

## Validation strategy

1. **Visual comparison**: `plot(fit)` + `text(fit)` shows the tree structure — count
   nodes and edges, verify they match our output
2. **`labels(fit, collapse=TRUE)`**: Compare our extracted split labels against rpart's
   own label generation — these are the ground truth since they handle `ncat` sign
3. **`path.rpart(fit)`**: Returns the path from root to each node — can verify edge
   sequences
4. **Stump test**: Fit with `maxdepth=0`, verify empty edgelist and single-node nodelist
5. **Categorical predictor test**: Verify csplit decoding produces correct level sets
6. **`ncat` sign test**: Verify our `split_op` extraction matches the label for both
   `ncat = -1` and `ncat = +1` cases

## Complexity assessment: LOW-MEDIUM (revised up from LOW)

The binary heap numbering makes parent-child extraction trivial — no traversal
algorithm needed. However:

- **The `ncat` sign gotcha** means we cannot assume left = `<` and right = `>=`. We
  must either parse labels or check the sign from the splits matrix. This is the kind
  of thing that's easy to get wrong and hard to notice in simple tests.
- Walking the `splits` matrix (sequential consumption with ncompete/nsurrogate offsets)
  requires careful bookkeeping.
- Categorical splits via `csplit` add another layer.

Using `labels(fit, collapse = TRUE)` as the source of truth for edge conditions
side-steps the `ncat` issue, but we still need splits-matrix walking for `split_point`.


---

# gbm

## Fit a model and inspect

We use `mtcars` for the primary model — it has 11 numeric variables and allows deeper
trees. We fit with `interaction.depth = 4` and 10 trees so we can see meaningful
structure and verify the missing-sentinel pattern at scale.

```{r gbm-fit}
library(gbm)
set.seed(42)
fit_gbm <- gbm(mpg ~ ., data = mtcars,
                distribution = "gaussian", n.trees = 10,
                interaction.depth = 4, shrinkage = 0.1,
                n.minobsinnode = 3, bag.fraction = 0.8)
cat("Number of trees:", fit_gbm$n.trees, "\n")
cat("Interaction depth:", fit_gbm$interaction.depth, "\n")
cat("num.classes:", fit_gbm$num.classes, "\n")
```

### `pretty.gbm.tree()` — the canonical tree accessor

```{r gbm-pretty}
pt <- pretty.gbm.tree(fit_gbm, i.tree = 1)
pt
```

### Internal list structure of a single tree

```{r gbm-tree-raw}
# trees[[i]] is an unnamed list of 8 vectors
cat("Tree 1 internal structure (list of 8 vectors):\n")
str(fit_gbm$trees[[1]])
```

### Node numbering and the 3d+1 formula

```{r gbm-nodes}
cat("Row names (0-based node IDs):", rownames(pt), "\n")
cat("Number of nodes:", nrow(pt), "\n")

# Identify internal vs terminal vs missing-sentinel
internal <- pt[pt$SplitVar != -1, ]
terminal <- pt[pt$SplitVar == -1, ]

cat("\nInternal nodes:", nrow(internal), "\n")
cat("Terminal nodes (SplitVar == -1):", nrow(terminal), "\n")

# Show the parent-child structure
if (nrow(internal) > 0) {
  cat("\nEdge structure from internal nodes:\n")
  edge_info <- data.frame(
    from = as.integer(rownames(internal)),
    left_child = internal$LeftNode,
    right_child = internal$RightNode,
    missing_child = internal$MissingNode,
    split_var_idx = internal$SplitVar,
    split_var_name = fit_gbm$var.names[internal$SplitVar + 1],
    split_threshold = round(internal$SplitCodePred, 4)
  )
  print(edge_info)
}
```

### THE CRITICAL GOTCHA: Missing-sentinel nodes

```{r gbm-missing-sentinel}
# gbm creates a SEPARATE physical node for each internal node's missing path
# These are NOT real tree nodes — they're routing destinations for NA values
# Formula: total rows = 3 * (num_internal) + 1

n_internal <- nrow(internal)
cat("Internal nodes:", n_internal, "\n")
cat("Expected total: 3 *", n_internal, "+ 1 =", 3 * n_internal + 1, "nodes\n")
cat("Actual:", nrow(pt), "nodes\n")

# Which nodes are missing-sentinels?
if (nrow(internal) > 0) {
  missing_ids <- internal$MissingNode
  left_ids <- internal$LeftNode
  right_ids <- internal$RightNode
  real_child_ids <- c(left_ids, right_ids)

  cat("\nMissing-sentinel node IDs:", sort(missing_ids), "\n")
  cat("Real child IDs (left+right):", sort(unique(real_child_ids)), "\n")
  cat("Are missing-sentinels disjoint from left/right children?",
      !any(missing_ids %in% real_child_ids), "\n")

  # Show which nodes are what
  all_ids <- 0:(nrow(pt) - 1)
  role <- ifelse(all_ids %in% as.integer(rownames(internal)), "internal",
          ifelse(all_ids %in% missing_ids, "MISSING-SENTINEL",
                 "leaf"))
  cat("\nNode roles:\n")
  print(data.frame(node = all_ids, role = role))
}
```

### Verify the pattern across multiple trees

```{r gbm-multi-tree}
cat("Tree structure summary across all", fit_gbm$n.trees, "trees:\n\n")
for (i in 1:fit_gbm$n.trees) {
  pt_i <- pretty.gbm.tree(fit_gbm, i.tree = i)
  n_int <- sum(pt_i$SplitVar != -1)
  n_total <- nrow(pt_i)
  formula_ok <- n_total == 3 * n_int + 1
  cat(sprintf("Tree %2d: %2d total, %2d internal, %2d terminal | 3d+1=%2d | match=%s\n",
              i, n_total, n_int, n_total - n_int,
              3 * n_int + 1, formula_ok))
}
```

### Variable metadata

```{r gbm-vars}
cat("Variable names (0-based index -> name):\n")
for (i in seq_along(fit_gbm$var.names)) {
  cat(sprintf("  [%d] %s (type=%d)\n", i - 1,
              fit_gbm$var.names[i], fit_gbm$var.type[i]))
}
```

### Reconstruct edges for Tree 1 — excluding missing-sentinels

```{r gbm-edges-demo}
pt1 <- pretty.gbm.tree(fit_gbm, i.tree = 1)
int1 <- pt1[pt1$SplitVar != -1, ]

if (nrow(int1) > 0) {
  edges_gbm <- data.frame(
    from = rep(as.integer(rownames(int1)), 2),
    to = c(int1$LeftNode, int1$RightNode),
    direction = c(rep("left", nrow(int1)), rep("right", nrow(int1))),
    stringsAsFactors = FALSE
  )
  real_nodes <- sort(unique(c(edges_gbm$from, edges_gbm$to)))

  cat("Edges for Tree 1:\n")
  print(edges_gbm)
  cat("\nReal tree nodes:", real_nodes, "\n")
  cat("Total real nodes:", length(real_nodes), "\n")
  cat("Expected (2*d + 1):", 2 * nrow(int1) + 1, "\n")

  # Show the missing-sentinel nodes we excluded
  all_node_ids <- 0:(nrow(pt1) - 1)
  excluded <- setdiff(all_node_ids, real_nodes)
  cat("Excluded missing-sentinel nodes:", excluded, "\n")
}
```

### c.splits — categorical split encoding

```{r gbm-csplits}
cat("c.splits length:", length(fit_gbm$c.splits), "\n")
if (length(fit_gbm$c.splits) > 0) {
  cat("First c.split entry:\n")
  print(fit_gbm$c.splits[[1]])
  cat("Values: 1=LEFT, 255=RIGHT, -1=missing\n")
} else {
  cat("(no categorical splits — mtcars predictors are all numeric)\n")
}
```

### Test with categorical predictors and deeper trees

```{r gbm-categorical}
set.seed(1)
n <- 300
df_cat_gbm <- data.frame(
  y = rnorm(n),
  x_cat = factor(sample(c("A","B","C","D","E"), n, replace = TRUE)),
  x_ord = factor(sample(c("low","med","high"), n, replace = TRUE),
                 levels = c("low","med","high"), ordered = TRUE),
  x_num1 = rnorm(n),
  x_num2 = rpois(n, 3)
)
# Make y depend on predictors
df_cat_gbm$y <- df_cat_gbm$y +
  ifelse(df_cat_gbm$x_cat %in% c("A","B"), 2, 0) +
  df_cat_gbm$x_num1 * 1.5

fit_gbm_cat <- gbm(y ~ ., data = df_cat_gbm, distribution = "gaussian",
                    n.trees = 10, interaction.depth = 4, shrinkage = 0.1,
                    n.minobsinnode = 5)
cat("var.names:", fit_gbm_cat$var.names, "\n")
cat("var.type:", fit_gbm_cat$var.type, "\n")
cat("c.splits length:", length(fit_gbm_cat$c.splits), "\n\n")

if (length(fit_gbm_cat$c.splits) > 0) {
  cat("c.splits entries (first 3):\n")
  for (j in seq_len(min(3, length(fit_gbm_cat$c.splits)))) {
    cat(sprintf("  c.splits[[%d]]: ", j))
    cat(fit_gbm_cat$c.splits[[j]], "\n")
  }
  cat("\nvar.levels:\n")
  str(fit_gbm_cat$var.levels)

  # Show a tree with a categorical split
  for (i in 1:10) {
    pt_i <- pretty.gbm.tree(fit_gbm_cat, i.tree = i)
    int_i <- pt_i[pt_i$SplitVar != -1, ]
    has_cat <- any(fit_gbm_cat$var.type[int_i$SplitVar + 1] > 0)
    if (has_cat) {
      cat(sprintf("\nTree %d has categorical split(s):\n", i))
      for (r in seq_len(nrow(int_i))) {
        vj <- int_i$SplitVar[r]
        if (fit_gbm_cat$var.type[vj + 1] > 0) {
          cs_idx <- as.integer(int_i$SplitCodePred[r])
          levels_j <- fit_gbm_cat$var.levels[[vj + 1]]
          cs_vec <- fit_gbm_cat$c.splits[[cs_idx + 1]]
          cat(sprintf("  Node %s: %s — left={%s} right={%s}\n",
                      rownames(int_i)[r],
                      fit_gbm_cat$var.names[vj + 1],
                      paste(levels_j[cs_vec == 1], collapse=","),
                      paste(levels_j[cs_vec != 1], collapse=",")))
        }
      }
      break
    }
  }
}
```

### Multinomial classification

```{r gbm-multinomial}
set.seed(42)
fit_gbm_multi <- gbm(Species ~ ., data = iris,
                      distribution = "multinomial", n.trees = 5,
                      interaction.depth = 3, shrinkage = 0.1,
                      n.minobsinnode = 5)
cat("num.classes:", fit_gbm_multi$num.classes, "\n")
cat("Total trees stored:", length(fit_gbm_multi$trees), "\n")
cat("Expected (n.trees * num.classes):", 5 * 3, "\n\n")

cat("Tree index -> (iteration, class) mapping:\n")
for (ti in 1:min(9, length(fit_gbm_multi$trees))) {
  iter <- ((ti - 1) %/% fit_gbm_multi$num.classes) + 1
  cls <- ((ti - 1) %% fit_gbm_multi$num.classes) + 1
  pt_i <- pretty.gbm.tree(fit_gbm_multi, i.tree = ti)
  cat(sprintf("  trees[[%2d]]: iteration=%d, class=%d (%s), nodes=%d\n",
              ti, iter, cls, levels(iris$Species)[cls], nrow(pt_i)))
}
```


## My interpretation of edges and nodes

**Nodes** = rows of `pretty.gbm.tree()` output, BUT with an important caveat:
**missing-sentinel nodes must be excluded** from the graph. They are phantom routing
nodes for NA handling, not real tree structure.

The real tree consists of:
- Internal nodes: `SplitVar != -1`
- Leaf nodes: `SplitVar == -1` AND reachable via LeftNode/RightNode (not MissingNode)

**Edges** = `LeftNode` and `RightNode` pointers from each internal node. Node IDs are
0-based integers, explicitly stored (unlike rpart's implicit heap numbering).

**Split semantics**: `x < threshold` goes LEFT, `x >= threshold` goes RIGHT.
For categoricals: index into global `c.splits` pool, where `1` = left, `255` = right.

## Approximate implementation approach

**`edgelist.gbm`:**

1. Call `pretty.gbm.tree(input_object, i.tree)` for selected tree(s)
2. Filter to internal nodes (`SplitVar != -1`)
3. Build edges: `from` = internal node ID (twice), `to` = c(LeftNode, RightNode)
4. Attach `split_var_name` via `var.names[SplitVar + 1]` (0-based indexing!)
5. Attach `split_point` from `SplitCodePred` (for continuous; for categorical, this is
   a csplit index — need to decode or store raw)
6. Support `treenum` parameter for filtering (handle `num.classes > 1` for multinomial)
7. **Exclude MissingNode edges** — they are not part of the tree topology

**`nodelist.gbm`:**

1. Get the set of node IDs that appear in edges (from + to), which automatically
   excludes missing-sentinel nodes
2. Join back to `pretty.gbm.tree` for attributes: Weight, Prediction, SplitVar, etc.
3. Include: name, is_leaf, split_var, split_var_name, split_point, prediction, weight,
   error_reduction, treenum

## Validation strategy

1. **Node count formula**: For `interaction.depth = d`, expect `d` internal nodes,
   `d + 1` leaves, `d` missing-sentinels, total = `3d + 1`. Verify our edgelist
   produces exactly `d` edges * 2 directions = `2d` edge rows, touching `2d + 1`
   unique nodes
2. **Manual trace**: For a 2-split tree (interaction.depth=2), manually trace the left/
   right edges and verify they form a valid binary tree
3. **Compare with `pretty.gbm.tree()` output directly** — our edges should cover exactly
   the LeftNode/RightNode columns
4. **Categorical validation**: Fit a model with known factor levels, decode c.splits, and
   verify the level assignments make sense against the data
5. **Stump test**: `interaction.depth = 1` gives 1 split, 2 leaves, 1 missing-sentinel
   (4 total rows). Verify our edgelist has exactly 2 edges

## Complexity assessment: MEDIUM-HIGH (flagged)

**Why this is flagged:**

1. **Missing-sentinel nodes** are a non-obvious structural feature. Every internal node
   creates a phantom "missing" child. If we naively include all `SplitVar == -1` nodes,
   we get extra nodes that don't belong in the tree. Must explicitly filter.

2. **0-based indexing everywhere** — `SplitVar`, `LeftNode`, `RightNode`, `c.splits`
   indices, row names. Easy off-by-one errors when converting to R's 1-based convention.

3. **`c.splits` is a global pool** across all trees. The index in `SplitCodePred` is a
   0-based global index. Must add 1 for R list indexing.

4. **Multinomial classification** stores `num.classes` trees per boosting iteration.
   The `treenum` parameter needs to account for this (tree index = `(iter-1)*K + class`).

5. **`Prediction` column semantics differ** between internal and terminal nodes — leaf
   predictions are the actual model contribution, internal node predictions are weighted
   averages (less useful).


---

# xgboost

## Fit a model and inspect

We use a synthetic 500-row dataset so xgboost has enough data to actually grow deep
trees (mtcars at 32 rows barely splits). `max_depth = 5` with 8 predictors gives us
trees with 30-60 nodes — enough structure to verify our understanding.

```{r xgb-fit}
library(xgboost)

# Synthetic dataset — 500 obs, 8 numeric predictors, non-linear response
set.seed(42)
n <- 500
X_syn <- matrix(rnorm(n * 8), nrow = n)
colnames(X_syn) <- paste0("x", 1:8)
y_syn <- 2*X_syn[,1] - 1.5*X_syn[,2] + X_syn[,3]*X_syn[,4] +
         0.5*X_syn[,5]^2 + rnorm(n, 0, 0.5)

fit_xgb <- xgboost(data = X_syn, label = y_syn,
                    objective = "reg:squarederror",
                    nrounds = 10, max_depth = 5,
                    eta = 0.3, verbose = 0,
                    min_child_weight = 3)
cat("nrounds:", fit_xgb$niter, "\n")
cat("feature_names:", fit_xgb$feature_names, "\n")
```

### `xgb.model.dt.tree()` — the key function

```{r xgb-dt-tree}
dt <- xgb.model.dt.tree(model = fit_xgb)
cat("Total rows (all trees, all nodes):", nrow(dt), "\n")
cat("Trees:", length(unique(dt$Tree)), "\n\n")
head(dt, 20)
```

### Column types and meanings

```{r xgb-columns}
str(dt)
```

### Per-tree summary

```{r xgb-per-tree}
for (t in unique(dt$Tree)) {
  sub <- dt[dt$Tree == t, ]
  n_split <- sum(sub$Feature != "Leaf")
  n_leaf <- sum(sub$Feature == "Leaf")
  cat(sprintf("Tree %2d: %2d nodes (%2d split, %2d leaf)\n",
              t, nrow(sub), n_split, n_leaf))
}
```

### Node ID format

```{r xgb-id-format}
cat("Sample IDs:", head(dt$ID, 15), "\n")
cat("Format: 'Tree-Node' (both 0-based)\n\n")

# Show structure of a single deeper tree
tree0 <- dt[dt$Tree == 0, ]
cat("Tree 0 — full structure:\n")
print(tree0[, c("Node", "ID", "Feature", "Split", "Yes", "No", "Missing",
                "Quality", "Cover")])
```

### Leaf node identification

```{r xgb-leaves}
leaves <- dt[dt$Feature == "Leaf", ]
splits <- dt[dt$Feature != "Leaf", ]

cat("Total nodes:", nrow(dt), "\n")
cat("Split (internal) nodes:", nrow(splits), "\n")
cat("Leaf nodes:", nrow(leaves), "\n")

cat("\nSample leaf row:\n")
print(leaves[1, ])
cat("\nSample split row:\n")
print(splits[1, ])
```

### Yes / No / Missing columns

```{r xgb-yes-no}
cat("Yes/No/Missing are ID strings pointing to child nodes:\n\n")
print(splits[1:5, c("ID", "Feature", "Split", "Yes", "No", "Missing")])

cat("\n\nMissing value routing across all split nodes:\n")
cat("Missing == Yes:", sum(splits$Missing == splits$Yes, na.rm = TRUE),
    "out of", nrow(splits), "\n")
cat("Missing == No:", sum(splits$Missing == splits$No, na.rm = TRUE),
    "out of", nrow(splits), "\n")
cat("Missing == neither:", sum(splits$Missing != splits$Yes &
    splits$Missing != splits$No, na.rm = TRUE), "\n")
```

### Edge construction — directly from split nodes

```{r xgb-edges}
# This is exactly how xgb.plot.tree does it internally
edges_xgb <- data.frame(
  from = c(splits$ID, splits$ID),
  to = c(splits$Yes, splits$No),
  branch = c(rep("yes", nrow(splits)), rep("no", nrow(splits))),
  feature = c(splits$Feature, splits$Feature),
  split = c(splits$Split, splits$Split),
  tree = c(splits$Tree, splits$Tree),
  stringsAsFactors = FALSE
)
cat("Total edge count:", nrow(edges_xgb), "\n")
cat("Unique from nodes:", length(unique(edges_xgb$from)), "\n")
cat("Unique to nodes:", length(unique(edges_xgb$to)), "\n\n")

# Show edges for Tree 0 only
cat("Edges for Tree 0:\n")
print(edges_xgb[edges_xgb$tree == 0, c("from","to","branch","feature","split")])
```

### Quality column dual meaning

```{r xgb-quality}
cat("Quality for split nodes = information gain:\n")
summary(splits$Quality)
cat("\nQuality for leaf nodes = leaf score (raw prediction):\n")
summary(leaves$Quality)
```

### Verify tree structure — every to-node should exist as a row

```{r xgb-verify}
# Confirm that every child ID referenced in Yes/No actually exists in the dt
all_ids <- dt$ID
yes_ids <- splits$Yes
no_ids <- splits$No

cat("All Yes-children exist in ID column:", all(yes_ids %in% all_ids), "\n")
cat("All No-children exist in ID column:", all(no_ids %in% all_ids), "\n")

# Confirm no phantom/sentinel nodes (unlike gbm)
cat("\nTotal unique nodes from edges:",
    length(unique(c(edges_xgb$from, edges_xgb$to))), "\n")
cat("Total rows in dt:", nrow(dt), "\n")
cat("These should be equal (no phantom nodes):",
    length(unique(c(edges_xgb$from, edges_xgb$to))) +
    1 == nrow(dt[dt$Tree == 0, ]),  # root is only in 'from', not 'to'... actually roots are in from
    "\n")

# Better check: for each tree, nodes = split_nodes + leaf_nodes, edges = 2*split_nodes
for (t in 0:2) {
  sub <- dt[dt$Tree == t, ]
  n_s <- sum(sub$Feature != "Leaf")
  n_l <- sum(sub$Feature == "Leaf")
  cat(sprintf("Tree %d: %d split + %d leaf = %d total, edges = %d (should be 2*%d = %d)\n",
              t, n_s, n_l, n_s + n_l, 2 * n_s, n_s, 2 * n_s))
}
```

### `xgb.dump()` — text representation for cross-validation

```{r xgb-dump}
dump_text <- xgb.dump(fit_xgb, with_stats = TRUE)
cat("First 30 lines of text dump:\n")
cat(paste(head(dump_text, 30), collapse = "\n"))
```

### Multi-class model

```{r xgb-multiclass}
X_iris <- as.matrix(iris[, 1:4])
colnames(X_iris) <- colnames(iris)[1:4]
y_iris <- as.integer(iris$Species) - 1L

set.seed(42)
fit_xgb_mc <- xgboost(data = X_iris, label = y_iris,
                       objective = "multi:softmax", num_class = 3,
                       nrounds = 5, max_depth = 4,
                       eta = 0.3, verbose = 0)

dt_mc <- xgb.model.dt.tree(model = fit_xgb_mc)
cat("Total trees:", length(unique(dt_mc$Tree)), "\n")
cat("Expected (3 classes * 5 rounds):", 3 * 5, "\n\n")

# Tree -> (class, round) mapping
cat("Tree -> (class, round) mapping:\n")
for (t in unique(dt_mc$Tree)) {
  cls <- t %% 3
  rnd <- t %/% 3
  sub <- dt_mc[dt_mc$Tree == t, ]
  cat(sprintf("  Tree %2d -> class %d (%s), round %d: %d nodes\n",
              t, cls, levels(iris$Species)[cls + 1], rnd, nrow(sub)))
}
```

### Feature names handling — with vs without

```{r xgb-feature-names}
cat("model$feature_names:", fit_xgb$feature_names, "\n")
cat("Feature column uses actual names (from colnames(X)):\n")
head(dt[dt$Feature != "Leaf", c("ID","Feature","Split")], 8)

# Compare with no feature names (unnamed matrix)
X_nonames <- unname(X_syn)
fit_xgb_nonames <- xgboost(data = X_nonames, label = y_syn,
                            objective = "reg:squarederror",
                            nrounds = 2, max_depth = 3,
                            eta = 0.3, verbose = 0)
dt_nonames <- xgb.model.dt.tree(model = fit_xgb_nonames)
cat("\nWithout feature_names (0-based indices as strings):\n")
head(dt_nonames[dt_nonames$Feature != "Leaf", c("ID","Feature","Split")], 6)
```

### use_int_id option

```{r xgb-int-id}
dt_int <- xgb.model.dt.tree(model = fit_xgb, use_int_id = TRUE)
cat("Columns with use_int_id=TRUE:\n")
cat(names(dt_int), "\n")
cat("\nNote: ID column is DROPPED; Yes/No/Missing become integers\n")
head(dt_int[dt_int$Feature != "Leaf",
            c("Tree", "Node", "Feature", "Yes", "No", "Missing")], 8)
```


## My interpretation of edges and nodes

**Nodes** = rows of `xgb.model.dt.tree()`. Each node has a globally unique `ID`
("Tree-Node" format). Leaf nodes have `Feature == "Leaf"`.

**Edges** = `Yes` and `No` pointers from each split node. These are ID strings,
so from/to values are already globally unique across trees. The `Missing` pointer
is a third routing path for NA values — typically equals `Yes` and can be annotated
on edges but shouldn't create additional edges.

**Split semantics**: `feature < Split` goes to `Yes` (left), `feature >= Split` goes
to `No` (right).

## Approximate implementation approach

**`edgelist.xgb.Booster`:**

1. Call `xgb.model.dt.tree(model = input_object)` — this gives everything we need
2. Filter to split nodes (`Feature != "Leaf"`)
3. Build edges: `from = ID` (twice), `to = c(Yes, No)`
4. Attach: feature, split threshold, gain (Quality), cover, tree number
5. Optionally annotate which branch the Missing path follows
6. Support `treenum` parameter via `trees` argument to `xgb.model.dt.tree()`
7. **Decision needed**: Use string IDs ("0-3") or integer node IDs? String IDs are
   globally unique out of the box; integer IDs need tree-prefixing (like randomForest)

**`nodelist.xgb.Booster`:**

1. All rows from `xgb.model.dt.tree()` — no phantom nodes to filter (unlike gbm)
2. Include: name (ID), tree, node, is_leaf, feature, split, quality/gain, cover,
   leaf_score (Quality for leaves), label
3. Feature names: handle both named and unnamed models (check `feature_names`)

**`as_igraph.xgb.Booster`:**

Should work naturally — string IDs in from/to give globally unique node names, so
`graph_from_data_frame()` handles multi-tree graphs as disconnected components.

## Validation strategy

1. **`xgb.plot.tree(model, trees = 0)`** draws the tree using DiagrammeR. Compare
   node count, edge count, and split labels against our output
2. **`xgb.dump(model, with_stats = TRUE)`** gives a text representation — verify that
   our edgelist captures the same structure
3. **Node count check**: For a single tree with `max_depth = d`, expect at most
   `2^(d+1) - 1` nodes. Verify our output is within bounds
4. **Round-trip test**: Build igraph from our edgelist/nodelist, verify it has the
   expected number of connected components (one per tree)
5. **Multi-class test**: Verify tree count = `num_class * nrounds` and that tree-to-class
   mapping is correct
6. **Feature name test**: Fit with and without `feature_names`, verify Feature column
   content differs appropriately

## Complexity assessment: LOW-MEDIUM

**Why relatively straightforward:**

- `xgb.model.dt.tree()` is a clean, well-structured data.table output — no sequential
  walking of splits matrices, no phantom nodes
- Yes/No/Missing are explicit ID references — no heap arithmetic or daughter columns
- String IDs are globally unique — no need for tree-prefixing logic
- The function handles all the internal parsing (xgb.dump text -> data.table)

**Watch out for:**

- **Feature names**: When `model$feature_names` is NULL, the Feature column contains
  0-based integer indices as strings (e.g., "2" not "Petal.Length"). Need to handle
  this gracefully — either pass feature_names explicitly or document the behavior
- **0-based tree indexing**: xgboost uses 0-based tree indices vs our package's
  convention of 1-based `treenum` (from randomForest). Need to convert
- **Multi-class tree ordering**: `Tree %% num_class` gives the class — need to decide
  if/how to expose this


---

# Comparison and implementation priority

```{r comparison-table, echo=FALSE}
comparison <- data.frame(
  Model = c("rpart", "xgboost", "gbm"),
  Complexity = c("Low-Medium", "Low-Medium", "Medium-High"),
  `Tree Access` = c(
    "rownames(frame) + heap math",
    "xgb.model.dt.tree() data.table",
    "pretty.gbm.tree() data.frame"
  ),
  `Edge Source` = c(
    "Implicit: parent = id %/% 2",
    "Explicit: Yes/No ID columns",
    "Explicit: LeftNode/RightNode columns"
  ),
  `Node IDs` = c(
    "1-based heap (gaps possible)",
    "String 'Tree-Node' (0-based)",
    "0-based integers"
  ),
  `Gotchas` = c(
    "ncat sign flips split direction; splits matrix sequential walk; csplit for categoricals",
    "0-based everything; feature_names may be NULL",
    "Missing-sentinel phantom nodes; 0-based indexing; global c.splits pool; multinomial K trees per round"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(comparison, caption = "Implementation comparison across stub models")
```

**Suggested implementation order:** rpart -> xgboost -> gbm

- **rpart first**: Simplest structure, already in base R, well-documented internals
- **xgboost second**: Clean API via `xgb.model.dt.tree()`, no phantom nodes
- **gbm last**: Most gotchas (missing-sentinels, global c.splits pool, 0-based
  everything, multinomial indexing)
