---
title: "Working with edgelists and nodelists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with edgelists and nodelists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
pkgload::load_all()
# library(networkformat)
```

`edgelist()` and `nodelist()` are the core functions of **networkformat**.
They extract network structure from tree models and tabular data as plain
data frames, giving you full control over filtering, transforming, and
passing the results to igraph or any other tool.

For quick visualization without touching the raw data, see
`vignette("visualization")`.

# Decision tree

## edgelist()

```{r tree-edges, eval = requireNamespace("tree", quietly = TRUE)}
tr <- tree::tree(Species ~ Sepal.Length + Sepal.Width, data = iris)
edges <- edgelist(tr)
edges
```

Columns:

| Column | Description |
|--------|-------------|
| `from` | Parent node index |
| `to` | Child node index |
| `label` | Full split condition (e.g. `"Sepal.Length <5.45"`) |
| `split_var` | Variable name |
| `split_op` | `"<"` or `">="` (NA for categorical splits) |
| `split_point` | Numeric threshold (NA for categorical) |

The parsed columns let you filter or restyle without regex:

```{r tree-filter, eval = requireNamespace("tree", quietly = TRUE)}
# All edges that split on Sepal.Length
edges[edges$split_var == "Sepal.Length", ]

# Thresholds above 5
edges[!is.na(edges$split_point) & edges$split_point > 5, ]
```

## nodelist()

```{r tree-nodes, eval = requireNamespace("tree", quietly = TRUE)}
nodes <- nodelist(tr)
nodes
```

Columns:

| Column | Description |
|--------|-------------|
| `node` | Node ID (matches edgelist `from`/`to`) |
| `var` | Split variable, or `"<leaf>"` |
| `n` | Observation count |
| `dev` | Deviance |
| `yval` | Predicted value |
| `is_leaf` | Logical |
| `label` | Display label: `"<var>\nn=<n>"` or `"<yval>\nn=<n>"` |

The `label` column is ready for plotting --- no manual `ifelse()` needed.

## Pairing edges and nodes for igraph

Node IDs line up, so you can pass them straight to
`igraph::graph_from_data_frame()`:

```{r tree-igraph, eval = requireNamespace("tree", quietly = TRUE) && requireNamespace("igraph", quietly = TRUE)}
library(igraph)
g <- graph_from_data_frame(
  edges, directed = TRUE,
  vertices = data.frame(name = nodes$node, nodes[-1])
)
cat("Nodes:", vcount(g), " Edges:", ecount(g), "\n")
```

Or use `as_igraph(tr)` for the same result in one step.

## Visualizing the edgelist pipeline

The full workflow --- `edgelist()` -> `nodelist()` -> `graph_from_data_frame()` -> ggraph --- gives you control at every step:

```{r tree-viz, eval = requireNamespace("tree", quietly = TRUE) && requireNamespace("igraph", quietly = TRUE) && requireNamespace("ggraph", quietly = TRUE), fig.width = 9, fig.height = 6, message = FALSE}
library(ggraph)
library(tidygraph)

tg <- as_tbl_graph(g)

ggraph(tg, layout = "tree") +
  geom_edge_diagonal(colour = "grey60", width = 0.5) +
  geom_node_label(
    aes(label = label, fill = ifelse(is_leaf, yval, var)),
    size = 2.5, colour = "white", fontface = "bold",
    label.padding = unit(0.25, "lines"), label.r = unit(0.2, "lines")) +
  scale_fill_brewer(palette = "Set2", name = NULL) +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "Iris tree (built from edgelist + nodelist)")
```

# Random forest

## edgelist()

```{r rf-edges, eval = requireNamespace("randomForest", quietly = TRUE), message = FALSE}
library(randomForest)
rf <- randomForest(Species ~ ., data = iris, ntree = 3, maxnodes = 5)

# All trees
rf_edges <- edgelist(rf)
head(rf_edges)
```

Columns:

| Column | Description |
|--------|-------------|
| `source` | Parent node index |
| `target` | Child node index |
| `split_var` | Numeric variable index |
| `split_point` | Split threshold |
| `prediction` | Predicted value at parent |
| `treenum` | Tree number |
| `split_var_name` | Human-readable variable name |

### The `treenum` argument

Extract specific trees without subsetting:

```{r rf-treenum, eval = requireNamespace("randomForest", quietly = TRUE)}
# Single tree
t1 <- edgelist(rf, treenum = 1)
nrow(t1)

# Trees 1 and 3
t13 <- edgelist(rf, treenum = c(1, 3))
table(t13$treenum)
```

## nodelist()

```{r rf-nodes, eval = requireNamespace("randomForest", quietly = TRUE)}
rf_nodes <- nodelist(rf, treenum = 1)
rf_nodes
```

Columns:

| Column | Description |
|--------|-------------|
| `node` | Node ID (matches edgelist `source`/`target`) |
| `is_leaf` | Logical |
| `split_var` | Numeric variable index (NA for leaves) |
| `split_var_name` | Variable name (NA for leaves) |
| `split_point` | Split threshold (NA for leaves) |
| `prediction` | Predicted value |
| `treenum` | Tree number |
| `label` | `split_var_name` for internal nodes, `prediction` for leaves |

# Data frame

## edgelist()

When your data is already tabular, `edgelist()` reshapes it into
source-target format. Columns are specified with
[tidyselect](https://tidyselect.r-lib.org/reference/language.html)
syntax --- bare names, strings, numbers, or helpers like `starts_with()`.

```{r df-basic}
courses

edgelist(courses, source_cols = course, target_cols = (prereq))
```

### Multiple target columns

When you pass multiple `target_cols`, each produces a block of edges.
The `target_col` column identifies which relationship each edge came from:

```{r df-multi}
edgelist(courses, source_cols = course,
         target_cols = c(prereq, prereq2, crosslist))
```

### `na.rm` --- handling missing values

By default, rows where source or target is NA are removed. Set
`na.rm = FALSE` to keep them:

```{r df-narm}
# Default: NAs removed
nrow(edgelist(courses, source_cols = course,
              target_cols = c(prereq, prereq2, crosslist)))

# Keep NAs
nrow(edgelist(courses, source_cols = course,
              target_cols = c(prereq, prereq2, crosslist),
              na.rm = FALSE))
```

### `attr_cols` --- selecting edge attributes

By default, all columns not used as source or target are carried along.
Use `attr_cols` to select specific columns, or `c()` for none:

```{r df-attrs}
# Only source, target, metadata
edgelist(courses, source_cols = course, target_cols = prereq,
         attr_cols = c())

# Specific columns
edgelist(courses, source_cols = course, target_cols = prereq,
         attr_cols = c(dept, credits))
```

### `symmetric_cols` --- marking undirected relationships

Some target columns represent symmetric relationships (crosslists,
co-authorships). Use `symmetric_cols` to add a `directed` column:

```{r df-sym}
edgelist(courses, source_cols = course,
         target_cols = c(prereq, prereq2, crosslist),
         attr_cols = c(),
         symmetric_cols = crosslist)
```

Prerequisite edges get `directed = TRUE`; crosslist edges get
`directed = FALSE`. This is useful for setting edge styles (arrows vs.
dashed lines) without manual tagging.

## nodelist()

For data frames, `nodelist()` simply reorders columns so the ID column
comes first --- convenient for `igraph::graph_from_data_frame()`:

```{r df-nodelist}
nodelist(courses, id_col = course)
```

## Note on tidygraph

tidygraph provides its own `as_tbl_graph.data.frame()` method that
auto-detects edge data frames (columns named `from`/`to` or
`source`/`target`). If your edgelist output uses those column names,
you can pass it directly:

```{r tidygraph-note, eval = FALSE}
edges <- edgelist(courses, source_cols = course, target_cols = prereq)
# Rename to from/to for tidygraph auto-detection
names(edges)[1:2] <- c("from", "to")
tg <- tidygraph::as_tbl_graph(edges)
```

networkformat does not add its own `as_tbl_graph.data.frame()` method
since tidygraph already covers this case.

## Putting it together: course network

```{r courses-plot, eval = requireNamespace("igraph", quietly = TRUE) && requireNamespace("ggraph", quietly = TRUE), fig.width = 10, fig.height = 7, message = FALSE}
library(igraph)
library(ggraph)
library(tidygraph)

all_edges <- edgelist(courses, source_cols = course,
                      target_cols = c(prereq, prereq2, crosslist),
                      attr_cols = c(),
                      symmetric_cols = crosslist)

# Reverse prereq arrows so they point from prereq → dependent course
all_edges[c("source", "target")] <- all_edges[c("target", "source")]

# Crosslists are symmetric --- keep one direction per pair
all_edges <- all_edges[!(all_edges$directed == FALSE &
                         as.character(all_edges$source) >=
                         as.character(all_edges$target)), ]

nodes <- nodelist(courses, id_col = course)
g <- graph_from_data_frame(all_edges, vertices = nodes)
tg <- as_tbl_graph(g)

dept_pal <- c(STAT = "#5B8DB8", MATH = "#E07B54", DATA = "#6BAF6B", CS = "#9B6BB0")

ggraph(tg, layout = "sugiyama") +
  geom_edge_diagonal(
    aes(filter = (directed == TRUE)),
    arrow = arrow(length = unit(2, "mm"), type = "closed"),
    end_cap = circle(8, "mm"), start_cap = circle(8, "mm"),
    edge_width = 0.5, colour = "grey45") +
  geom_edge_arc(
    aes(filter = (directed == FALSE)),
    end_cap = circle(6, "mm"), start_cap = circle(6, "mm"),
    edge_width = 0.8, colour = "#9B59B6", linetype = "dashed",
    strength = 0.3) +
  geom_node_label(
    aes(label = name, fill = dept),
    colour = "white", fontface = "bold", size = 3,
    label.padding = unit(0.3, "lines"), label.r = unit(0.2, "lines")) +
  scale_fill_manual(values = dept_pal, name = "Department") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "Course prerequisite network",
       subtitle = "Solid arrows = prerequisites  |  Dashed curves = crosslists")
```

### Graph statistics

Once the data is in igraph, you can compute standard graph metrics:

```{r courses-stats, eval = requireNamespace("igraph", quietly = TRUE)}
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Components:", count_components(g), "\n")

# In-degree: how many prerequisites each course has
deg_in <- degree(g, mode = "in")
cat("\nMost prerequisites:\n")
print(sort(deg_in[deg_in > 0], decreasing = TRUE))

# Out-degree: how many courses depend on this one
deg_out <- degree(g, mode = "out")
cat("\nMost depended-on:\n")
print(sort(deg_out[deg_out > 0], decreasing = TRUE))
```

# Output reference

## edgelist()

| Input class | Columns returned |
|-------------|-----------------|
| `data.frame` | `source`, `target`, `source_col`, `target_col`, [`directed`], `<attr_cols>` |
| `tree` | `from`, `to`, `label`, `split_var`, `split_op`, `split_point` |
| `randomForest` | `source`, `target`, `split_var`, `split_point`, `prediction`, `treenum`, `split_var_name` |

## nodelist()

| Input class | Columns returned |
|-------------|-----------------|
| `data.frame` | Reordered input (ID column first) |
| `tree` | `node`, `var`, `n`, `dev`, `yval`, `is_leaf`, `label` |
| `randomForest` | `node`, `is_leaf`, `split_var`, `split_var_name`, `split_point`, `prediction`, `treenum`, `label` |

## as_igraph() / as_tbl_graph()

| Input class | `treenum` | Returns |
|-------------|-----------|---------|
| `tree` | --- | single graph |
| `randomForest` | `1` (default) | single graph (one tree) |
| `randomForest` | `NULL` | single graph (all trees as disconnected components) |
