---
title: "networkformat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{networkformat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Why networkformat?

Decision trees, random forests, and other tree-based models are networks ---
nodes connected by edges --- but they don't come in network format. The
output of `randomForest::getTree()` is a matrix of daughter indices and split
points. The `tree` package stores its structure in a frame with implicit
parent-child relationships. To visualize or analyze these models as graphs,
you have to write bespoke reshaping code every time.

**networkformat** does that reshaping for you. It extracts standard
**edgelists** (source-target pairs) and **nodelists** (node attributes) from
models and tabular data, putting everything in the format that the rest of
the R network ecosystem expects.

# Quick example

Five lines from model to plot:

```{r quick-example, eval = requireNamespace("tree", quietly = TRUE) && requireNamespace("ggraph", quietly = TRUE), fig.width = 9, fig.height = 6, message = FALSE}
pkgload::load_all()
# library(networkformat)
library(tidygraph)
library(ggraph)

tr <- tree::tree(Species ~ Sepal.Length + Sepal.Width, data = iris)
tg <- as_tbl_graph(tr)

ggraph(tg, layout = "tree") +
  geom_edge_diagonal(colour = "grey60") +
  geom_node_label(aes(label = label, fill = ifelse(is_leaf, yval, var)),
                  size = 2.5, colour = "white", fontface = "bold") +
  scale_fill_brewer(palette = "Set2", name = NULL) +
  theme_graph(base_family = "sans") +
  labs(title = "Iris classification tree")
```

# The R network ecosystem

Once your data is in edgelist/nodelist form, a large ecosystem of packages
is available, including:

**igraph** --- the workhorse graph library for R. Handles graph construction,
shortest paths, centrality, community detection, clustering, and much more.
Also converts to **adjacency matrices** (`as_adjacency_matrix()`) and
**incidence matrices** (`as_incidence_matrix()`) for linear algebra
approaches.

**tidygraph** --- a tidy interface built on top of igraph. Lets you
manipulate graphs with dplyr verbs (`mutate`, `filter`, `arrange`) while
keeping the underlying igraph structure. If you're comfortable with the
tidyverse, this is the natural way to work with graphs. Note that tidygraph
provides `as_tbl_graph.data.frame()`, which auto-detects edge data frames
with `from`/`to` or `source`/`target` columns --- so once you have an
edgelist from networkformat, you can pass it directly to tidygraph without
building an igraph first.

**ggraph** --- a ggplot2 extension for graph visualization, designed to work
with tidygraph/igraph objects. Supports tree layouts, force-directed layouts,
circular layouts, and more.

**network** / **statnet** --- an alternative graph ecosystem focused on
statistical network modeling (ERGMs, latent space models). igraph objects can
be converted to network objects via `intergraph::asNetwork()`.

networkformat gets your data into this ecosystem. From there, you can
visualize, compute statistics, fit models, or convert to matrices ---
whatever your analysis requires.

# What's implemented

## Currently available

| Model / input | `edgelist()` | `nodelist()` | `as_igraph()` |
|---------------|:---:|:---:|:---:|
| `tree` | yes | yes | yes |
| `randomForest` | yes | yes | yes |
| `data.frame` | yes | yes | --- |

For data frames, tidygraph's `as_tbl_graph()` already handles the
conversion from edge data frames to graph objects, so networkformat does
not duplicate that method.

## In progress

| Model | Status |
|-------|--------|
| `rpart` | Stub |
| `xgboost` (`xgb.Booster`) | Stub |
| `gbm` | Stub |

# Getting started

For visualizing tree models quickly, see `vignette("visualization")`.

For working directly with edgelists and nodelists --- including all the
available arguments and the data frame workflow --- see
`vignette("edgelist-nodelist")`.
