---
title: "Visualizing models as networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing models as networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
pkgload::load_all()
# library(networkformat)
library(ggraph)
library(tidygraph)
```

**networkformat** converts tree-based models into graph objects for
visualization with ggraph. The main entry points are:

- `as_igraph(x)` --- convert a model to an igraph object
- `as_tbl_graph(x)` --- convert to a tidygraph tbl_graph (for use with ggraph)

If you want to work with the raw edgelists and nodelists directly ---
filtering edges, computing graph statistics, or building custom igraph
objects --- see `vignette("edgelist-nodelist")`.

# Decision tree

```{r tree-fit, eval = requireNamespace("tree", quietly = TRUE)}
tr <- tree::tree(Species ~ Sepal.Length + Sepal.Width, data = iris)
tg <- as_tbl_graph(tr)
```

The graph carries all the attributes you need for plotting --- node labels,
split variables, leaf predictions, observation counts, and parsed edge
thresholds.

```{r tree-stats, eval = requireNamespace("tree", quietly = TRUE) && requireNamespace("igraph", quietly = TRUE)}
cat("Nodes:", igraph::vcount(tg), " Edges:", igraph::ecount(tg),
    " Depth:", max(igraph::distances(tg, mode = "out")[1, ]), "\n")
```

```{r tree-plot, eval = requireNamespace("tree", quietly = TRUE) && requireNamespace("ggraph", quietly = TRUE), fig.width = 10, fig.height = 7}
# Add a fill column: split variable for internal nodes, predicted class for leaves
tg <- tg %>%
  mutate(fill_var = ifelse(is_leaf, yval, var))

# Build a palette: one color per variable + one per class
node_df <- as.data.frame(tg, what = "vertices")
vars <- setdiff(unique(node_df$var), "<leaf>")
classes <- unique(node_df$yval[node_df$is_leaf])
var_pal <- c(
  setNames(c("#3B8EA5", "#D4A373"), vars[seq_len(min(length(vars), 2))]),
  setNames(c("#66c2a5", "#fc8d62", "#8da0cb"), classes[seq_len(min(length(classes), 3))])
)

ggraph(tg, layout = "tree") +
  geom_edge_diagonal(colour = "grey60", width = 0.5) +
  geom_edge_diagonal(
    aes(label = ifelse(is.na(split_point), "", paste0(split_op, split_point))),
    angle_calc = "along", label_dodge = unit(3, "mm"),
    label_size = 2.5, label_colour = "grey30", colour = NA) +
  geom_node_label(
    aes(label = label, fill = fill_var),
    size = 2.5, colour = "white", fontface = "bold",
    label.padding = unit(0.25, "lines"), label.r = unit(0.2, "lines")) +
  scale_fill_manual(values = var_pal, name = NULL) +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "Iris classification tree",
       subtitle = "Split on Sepal.Length and Sepal.Width")
```

Internal nodes are colored by the split variable. Leaf nodes show the
predicted class. Edge labels show the split condition (e.g. `<5.45`).

# Random forest

`as_tbl_graph(rf, treenum = NULL)` combines all trees into a single graph
with disconnected components --- one per tree. Each node carries a `treenum`
attribute so you can distinguish them.

```{r rf-fit, eval = requireNamespace("randomForest", quietly = TRUE), message = FALSE}
library(randomForest)
rf <- randomForest(Species ~ ., data = iris, ntree = 3, maxnodes = 5)

# All 3 trees in one graph
tg_rf <- as_tbl_graph(rf, treenum = NULL)
```

```{r rf-stats, eval = requireNamespace("randomForest", quietly = TRUE) && requireNamespace("igraph", quietly = TRUE)}
cat("Nodes:", igraph::vcount(tg_rf), " Edges:", igraph::ecount(tg_rf),
    " Components:", igraph::count_components(tg_rf), "\n")
```

```{r rf-plot, eval = requireNamespace("randomForest", quietly = TRUE) && requireNamespace("ggraph", quietly = TRUE), fig.width = 14, fig.height = 7}
# Add display columns directly to the graph
tg_rf <- tg_rf %>%
  mutate(
    fill_var = ifelse(is_leaf, "leaf", split_var_name),
    display_label = ifelse(
      is_leaf, label,
      paste0(split_var_name, "\n< ", round(split_point, 1))
    )
  )

# Palette: one color per predictor variable + muted for leaves
node_df <- as.data.frame(tg_rf, what = "vertices")
predictor_vars <- sort(unique(na.omit(node_df$split_var_name)))
var_colors <- c("#3B8EA5", "#D4A373", "#7B68AE", "#E07B54")
var_pal_rf <- c(
  setNames(var_colors[seq_along(predictor_vars)], predictor_vars),
  leaf = "#8fbc8f"
)

ggraph(tg_rf, layout = "tree") +
  geom_edge_link(
    arrow = arrow(length = unit(1.5, "mm"), type = "closed"),
    end_cap = circle(5, "mm"),
    colour = "grey70", width = 0.3) +
  geom_node_label(
    aes(label = display_label, fill = fill_var),
    size = 2.5, colour = "black", fontface = "bold",
    label.padding = unit(0.2, "lines"), label.r = unit(0.15, "lines"),
    alpha = 0.85) +
  scale_fill_manual(values = var_pal_rf, name = "Split variable") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "Random forest (3 trees)",
       subtitle = "Each disconnected component is one tree")
```

All three trees are visible in a single plot. Nodes are colored by the
split variable they use; leaf nodes are green. Internal node labels show the
variable and threshold.

## Single tree from a forest

For a closer look at one tree:

```{r rf-single, eval = requireNamespace("randomForest", quietly = TRUE) && requireNamespace("ggraph", quietly = TRUE), fig.width = 8, fig.height = 5}
tg1 <- as_tbl_graph(rf, treenum = 1) %>%
  mutate(
    fill_var = ifelse(is_leaf, "leaf", split_var_name),
    display_label = ifelse(
      is_leaf, label,
      paste0(split_var_name, "\n< ", round(split_point, 1))
    )
  )

ggraph(tg1, layout = "dendrogram") +
  geom_edge_elbow(colour = "grey65", width = 0.4) +
  geom_node_label(
    aes(label = display_label, fill = fill_var),
    colour = "black", size = 2.5, fontface = "bold",
    label.padding = unit(0.25, "lines"), label.r = unit(0.2, "lines"),
    label.size = 0.4, alpha = 0.2) +
  geom_node_text(
    aes(label = display_label),
    colour = "black", size = 2.5, fontface = "bold") +
  scale_fill_manual(values = var_pal_rf, name = "Split variable") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "Tree 1 from the random forest")
```

# Course prerequisite network

networkformat also handles tabular data. The bundled `courses` dataset
contains `r nrow(courses)` courses across `r length(unique(courses$dept))`
departments, with prerequisite and crosslist relationships.

```{r courses-net, eval = requireNamespace("igraph", quietly = TRUE) && requireNamespace("ggraph", quietly = TRUE), fig.width = 10, fig.height = 7, message = FALSE}
library(igraph)

all_edges <- edgelist(courses, source_cols = course,
                      target_cols = c(prereq, prereq2, crosslist),
                      attr_cols = c(),
                      symmetric_cols = crosslist)

# Reverse prereq arrows so they point from prereq → dependent course
all_edges[c("source", "target")] <- all_edges[c("target", "source")]

# Crosslists are symmetric --- keep one direction per pair
all_edges <- all_edges[!(all_edges$directed == FALSE &
                         as.character(all_edges$source) >=
                         as.character(all_edges$target)), ]

nodes <- nodelist(courses, id_col = course)
g <- graph_from_data_frame(all_edges, vertices = nodes)
tg_courses <- as_tbl_graph(g)

cat("Nodes:", vcount(g), " Edges:", ecount(g), "\n")

dept_pal <- c(STAT = "#5B8DB8", MATH = "#E07B54", DATA = "#6BAF6B", CS = "#9B6BB0")

ggraph(tg_courses, layout = "sugiyama") +
  geom_edge_diagonal(
    aes(filter = (directed == TRUE)),
    arrow = arrow(length = unit(2, "mm"), type = "closed"),
    end_cap = circle(8, "mm"), start_cap = circle(8, "mm"),
    edge_width = 0.5, colour = "grey45") +
  geom_edge_arc(
    aes(filter = (directed == FALSE)),
    end_cap = circle(6, "mm"), start_cap = circle(6, "mm"),
    edge_width = 0.8, colour = "#9B59B6", linetype = "dashed",
    strength = 0.3) +
  geom_node_label(
    aes(label = name, fill = dept),
    colour = "white", fontface = "bold", size = 3,
    label.padding = unit(0.3, "lines"), label.r = unit(0.2, "lines")) +
  scale_fill_manual(values = dept_pal, name = "Department") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "Course prerequisite network",
       subtitle = "Solid arrows = prerequisites  |  Dashed curves = crosslists")
```

For the full data frame workflow --- column selection, `na.rm`, `attr_cols`,
`symmetric_cols`, and graph statistics --- see `vignette("edgelist-nodelist")`.
