---
title: "Complete method reference"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(networkformat)
```

This article demonstrates every `edgelist()`, `nodelist()`, `as.igraph()`, and `as_tbl_graph()` method in the package. For a gentler introduction, see `vignette("networkformat")`.

## Atomic vectors

Vectors produce sequential edges: element *i* connects to element *i + 1*.

### edgelist

```{r}
edgelist(c("A", "B", "C", "D"))
```

Numeric vectors work the same way:

```{r}
edgelist(c(1, 2, 3))
```

Duplicate edges can be collapsed with `weights = TRUE`:

```{r}
edgelist(c("A", "B", "A", "B", "C"), weights = TRUE)
```

### nodelist

Returns unique values with frequency counts:

```{r}
nodelist(c("A", "B", "A", "B", "C"))
```

---

## Lists

Lists produce recursive parent-child edges. Each element becomes a node with a path-style ID.

### edgelist

```{r}
edgelist(list(a = list(b = 1, c = 2), d = 3))
```

Unnamed elements use positional indices:

```{r}
edgelist(list(1, 2, list(3, 4)))
```

Limit depth (root = 0, children = 1, ...):

```{r}
edgelist(list(a = list(b = list(c = 1))), max_depth = 2)
```

### nodelist

```{r}
nodelist(list(a = list(b = 1, c = 2), d = 3))
```

### S3 object fallthrough

S3 objects without a dedicated method are decomposed as plain lists:

```{r eval = FALSE}
fit <- lm(Sepal.Length ~ Sepal.Width, data = iris)
edgelist(fit)
#> No edgelist method for class 'lm'; treating as a plain list.
```

---

## Data frames

Data frame methods use tidyselect to specify which columns form edges.

### edgelist

**Basic usage** with the bundled `courses` dataset:

```{r}
data(courses)
edgelist(courses, source_cols = prereq, target_cols = course)
```

**Multiple source columns** — each source column produces a block of edges, identified by `from_col`:

```{r}
edgelist(courses, source_cols = c(prereq, prereq2), target_cols = course)
```

**Keeping NAs** (default is `na.rm = TRUE`):

```{r}
edgelist(courses, source_cols = prereq, target_cols = course, na.rm = FALSE)
```

**Controlling attribute columns** — by default all non-edge columns are carried through. Use `attr_cols` to select specific ones:

```{r}
edgelist(courses, source_cols = prereq, target_cols = course,
         attr_cols = c(dept, credits))
```

Or pass `c()` to drop all attributes:

```{r}
edgelist(courses, source_cols = prereq, target_cols = course, attr_cols = c())
```

**Symmetric (undirected) edges** with automatic deduplication:

```{r}
edgelist(courses, source_cols = course, target_cols = crosslist,
         symmetric_cols = crosslist, dedupe = TRUE)
```

**Edge weights** — collapse identical rows and count:

```{r}
edgelist(courses, source_cols = prereq, target_cols = course, weights = TRUE)
```

**Mixed directed and undirected edges:**

```{r}
edgelist(courses,
         source_cols = course,
         target_cols = c(prereq, crosslist),
         symmetric_cols = crosslist)
```

### nodelist

Reorders the data frame so the ID column comes first:

```{r}
nodelist(courses, id_col = course)
```

---

## Decision trees (tree)

```{r}
library(tree)
tr <- tree(Species ~ ., iris)
```

### edgelist

```{r}
edgelist(tr)
```

### nodelist

```{r}
nodelist(tr)
```

### as.igraph

```{r}
library(igraph)
g <- as.igraph(tr)
g
```

```{r}
vcount(g)
ecount(g)
```

### as_tbl_graph

```{r}
library(tidygraph)
as_tbl_graph(tr)
```

---

## Random forests (randomForest)

```{r}
library(randomForest)
set.seed(12)
rf <- randomForest(Species ~ ., iris, ntree = 3)
```

### edgelist

All trees:

```{r}
edgelist(rf)
```

A single tree:

```{r}
edgelist(rf, treenum = 1)
```

Multiple specific trees:

```{r}
edgelist(rf, treenum = c(1, 3))
```

### nodelist

```{r}
nodelist(rf, treenum = 1)
```

### as.igraph

Single tree:

```{r}
g <- as.igraph(rf, treenum = 1)
g
```

Multiple trees produce disconnected components:

```{r}
g_all <- as.igraph(rf)
components(g_all)$no
```

### as_tbl_graph

```{r}
as_tbl_graph(rf, treenum = 1)
```

---

## Recursive partitioning (rpart)

```{r}
library(rpart)
rp <- rpart(Species ~ ., iris)
```

### edgelist

```{r}
edgelist(rp)
```

### nodelist

```{r}
nodelist(rp)
```

### as.igraph

```{r}
g <- as.igraph(rp)
g
```

### as_tbl_graph

```{r}
as_tbl_graph(rp)
```

---

## Gradient boosted models (gbm)

```{r}
library(gbm)
set.seed(12)
gb <- gbm(as.numeric(Species == "setosa") ~ ., data = iris,
           distribution = "bernoulli", n.trees = 3, interaction.depth = 3)
```

### edgelist

All trees:

```{r}
edgelist(gb)
```

A single tree:

```{r}
edgelist(gb, treenum = 1)
```

### nodelist

```{r}
nodelist(gb, treenum = 1)
```

### as.igraph

```{r}
g <- as.igraph(gb, treenum = 1)
g
```

### as_tbl_graph

```{r}
as_tbl_graph(gb, treenum = 1)
```

---

## XGBoost (xgb.Booster)

```{r}
library(xgboost)
set.seed(12)
xg <- xgboost(
  x = as.matrix(iris[, 1:4]),
  y = iris$Species,
  max_depth = 3, nrounds = 3, nthreads = 1
)
```

### edgelist

All trees:

```{r}
edgelist(xg)
```

A single tree:

```{r}
edgelist(xg, treenum = 1)
```

### nodelist

```{r}
nodelist(xg, treenum = 1)
```

### as.igraph

```{r}
g <- as.igraph(xg, treenum = 1)
g
```

### as_tbl_graph

```{r}
as_tbl_graph(xg, treenum = 1)
```

---

## Output column reference

### edgelist columns by input type

| Input | Columns |
|-------|---------|
| vector | `from`, `to`, [`weight`] |
| list | `from`, `to`, `depth` |
| data.frame | `from`, `to`, `from_col`, `to_col`, [`directed`], [`weight`], \<attrs\> |
| tree | `from`, `to`, `label`, `split_var`, `split_op`, `split_point` |
| rpart | `from`, `to`, `label`, `split_var`, `split_op`, `split_point` |
| randomForest | `from`, `to`, `split_var`, `split_point`, `prediction`, `treenum`, `split_var_name` |
| gbm | `from`, `to`, `split_var`, `split_point`, `prediction`, `treenum`, `split_var_name` |
| xgb.Booster | `from`, `to`, `feature`, `split`, `quality`, `cover`, `treenum` |

### nodelist columns by input type

| Input | Columns |
|-------|---------|
| vector | `name`, `n` |
| list | `name`, `depth`, `type`, `n_children`, `label` |
| data.frame | (input columns reordered, `id_col` first) |
| tree | `name`, `var`, `n`, `dev`, `yval`, `is_leaf`, `label` |
| rpart | `name`, `var`, `n`, `dev`, `yval`, `is_leaf`, `label` |
| randomForest | `name`, `is_leaf`, `split_var`, `split_var_name`, `split_point`, `prediction`, `treenum`, `label` |
| gbm | `name`, `is_leaf`, `split_var`, `split_var_name`, `split_point`, `prediction`, `treenum`, `label` |
| xgb.Booster | `name`, `is_leaf`, `feature`, `split`, `quality`, `cover`, `treenum`, `label` |
